---
title: "MUSA 508 Assignment 1"
author: "Palak Agarwal & Shivani Rai"
date: "September 25th, 2020"
output:
  html_document:
  toc: true
toc_float: true
---

# Introduction 

This assignment looked at the **Bay Area Rapid Transit** and its analyzed all the 5 counties that it runs through to check if it has an effect on them. The main reason for choosing the BART system over the SF MTA system was to check if that cross county connection is more effective. SF being a big job center, attracts a lot of people from the surrounding counties in every day and the lack of vacancy and high housing values in the city forces people to stay in areas close to SF and commute everyday. The BART also connects other job centers in the bay area like San Jose, Oakland and the two airports.

```{r setup, include=FALSE}
#importing the library

library(rjson)
library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)
library(rlist)
options(scipen=999)
options(tigris_class = "sf")

```

## Load Themes

Create our mapping and plotting themes, color palettes and quantile break.

```{r load_themes, message = FALSE}
mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 24,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2)
  )
}

plotTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 24,colour = "black"),
    plot.subtitle = element_text(face="italic"),
    plot.caption = element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey60", size = 0.1),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.background = element_rect(fill = "grey60", color = "white"),
    strip.text = element_text(size=12),
    axis.title = element_text(size=12),
    axis.text = element_text(size=10),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"),
    strip.text.x = element_text(size = 14)
  )
}

# Load Quantile break functions

qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],0),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]]), digits = 3),
                 c(.01,.2,.4,.6,.8), na.rm=T)
  }
}

q5 <- function(variable) {as.factor(ntile(variable, 5))}

# Load hexadecimal color palette

palette5 <- c("#f6d908","#d6c054","#878dac","#3574e2","#005fc4","#004b9a","#b2a681")

```

## Next lets load the data we going to use

```{r load_data, message = FALSE}

#data import - bart lines and crime data 

sfo <- read.csv('C:/Users/agarw/Documents/MUSA508/MUSA-508-Assignment-1/Data/final_sfo_lines.csv')
```

## Loading the census API key 

```{r load_api_key, message = FALSE}
census_api_key("aea3dee2d96acb5101e94f3dcfa1b575f73d093a", overwrite = TRUE)
```

## Data wrangling with census data
### Year 2009 and 2017

We use year 2009 and 2017 in this report for our analyzes. The **state code** for California is **6** and we pass the 5 counties we working with as a list in the county. The counties with their county codes are :     
* Alameda        - 001   
* Contra Costa   - 013   
* San Franscisco - 075   
* San Mateo      - 081    
* Santa Clara    - 085    

```{r load_data for year 2009 and 2017, message = FALSE, warning=FALSE, results=FALSE}
# ---- Year 2009 tracts -----

tracts09 <-  
  get_acs(geography = "tract", variables = c("B25026_001E","B02001_002E","B19013_001E",
                                             "B25058_001E","B01001_006E","B01001_007E",
                                             "B01001_008E","B01001_009E","B01001_010E",
                                             "B01001_011E","B01001_012E","B01001_013E",
                                             "B01001_014E","B01001_015E","B01001_016E",
                                             "B01001_030E","B01001_031E","B01001_032E",
                                             "B01001_033E","B01001_034E","B01001_035E",
                                             "B01001_036E","B01001_037E","B01001_038E",
                                             "B01001_039E","B01001_040E","B12001_004E",
                                             "B12001_013E","B02001_003E","B02001_004E",
                                             "B02001_005E","B02001_006E","B03001_003E"), 
          year=2009, state=06, county=c(001,075,081,013,085), geometry=T) %>% 
  st_transform('ESRI:102241')

tracts09 <- 
  tracts09 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(variable, estimate) %>%
  dplyr::select(-geometry) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         Blacks = B02001_003,
         AmInd = B02001_004,
         Asian = B02001_005,
         Hawaiian = B02001_006,
         Hispanic = B03001_003,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         WorkingClassM1517 = B01001_006,
         WorkingClassM1819 = B01001_007,
         WorkingClassM20 = B01001_008,
         WorkingClassM21 = B01001_009,
         WorkingClassM2224 = B01001_010,
         WorkingClassM2529 = B01001_011,
         WorkingClassM3034 = B01001_012,
         WorkingClassM3539 = B01001_013,
         WorkingClassM4044 = B01001_014,
         WorkingClassM4549 = B01001_015,
         WorkingClassM5054 = B01001_016,
         WorkingClassF1517 = B01001_030,
         WorkingClassF1819 = B01001_031,
         WorkingClassF20 = B01001_032,
         WorkingClassF21 = B01001_033,
         WorkingClassF2224 = B01001_034,
         WorkingClassF2529 = B01001_035,
         WorkingClassF3034 = B01001_036,
         WorkingClassF3539 = B01001_037,
         WorkingClassF4044 = B01001_038,
         WorkingClassF4549 = B01001_039,
         WorkingClassF5054 = B01001_040,
         MaleMarried = B12001_004,
         FemaleMarried = B12001_013
         )

tracts209 <- 
  tracts09 %>%
  mutate(pctWorking = ifelse(TotalPop > 0, ((WorkingClassM1517 + WorkingClassM1819 
                                             + WorkingClassM20 + WorkingClassM21 
                                             + WorkingClassM2224 + WorkingClassM2529 
                                             + WorkingClassM3034 + WorkingClassM3539 
                                             + WorkingClassM4044 + WorkingClassM4549 
                                             + WorkingClassM5054 + WorkingClassF1517 
                                             + WorkingClassF1819 + WorkingClassF20 
                                             + WorkingClassF21   + WorkingClassF2224 
                                             + WorkingClassF2529 + WorkingClassF3034 
                                             + WorkingClassF3539 + WorkingClassF4044 
                                             + WorkingClassF4549 + WorkingClassF5054)
                                            /TotalPop * 100), 0),
         pctMarried = ifelse(TotalPop > 0, ((FemaleMarried + MaleMarried) / TotalPop * 100), 0),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop * 100, 0),
         pctOtherRace = ifelse(TotalPop > 0, ((Blacks + AmInd +Asian
                                               + Hawaiian + Hispanic) / TotalPop * 100), 0),
         
         year = "2009") %>%
  dplyr::select(-Whites,-FemaleMarried,-MaleMarried,-Blacks,-AmInd,-Asian,-Hawaiian,
                -Hispanic, -WorkingClassM1517,-WorkingClassM1819,-WorkingClassM20,
                -WorkingClassM21,-WorkingClassM2224,-WorkingClassM2529, 
                -WorkingClassM3034,-WorkingClassM3539,-WorkingClassM4044,
                -WorkingClassM4549,-WorkingClassM5054,-WorkingClassF1517, 
                -WorkingClassF1819,-WorkingClassF20,-WorkingClassF21,
                -WorkingClassF2224,-WorkingClassF2529,-WorkingClassF3034,
                -WorkingClassF3539,-WorkingClassF4044,-WorkingClassF4549,
                -WorkingClassF5054)

#----- Finding max values of popluation and rent to normalise the data 
# as the tract is not truly representative of the data near the station------

maxtotpop_2009 <- max(tracts209$TotalPop)
maxrent_2009 <- max(tracts209$MedRent , na.rm = T)
tracts2009 <- 
  tracts209 %>%
  mutate( normpop = TotalPop/maxtotpop_2009,
          normrent = MedRent/maxrent_2009)

# ---- Year 2017 tracts -----

tracts217 <-  
  get_acs(geography = "tract", variables = c("B25026_001E","B02001_002E","B19013_001E",
                                             "B25058_001E","B01001_006E","B01001_007E",
                                             "B01001_008E","B01001_009E","B01001_010E",
                                             "B01001_011E","B01001_012E","B01001_013E",
                                             "B01001_014E","B01001_015E","B01001_016E",
                                             "B01001_030E","B01001_031E","B01001_032E",
                                             "B01001_033E","B01001_034E","B01001_035E",
                                             "B01001_036E","B01001_037E","B01001_038E",
                                             "B01001_039E","B01001_040E","B12001_004E",
                                             "B12001_013E","B02001_003E","B02001_004E",
                                             "B02001_005E","B02001_006E","B03001_003E"), 
          year=2017, state=06, county=c(001,075,081,013,085), geometry=T) %>% 
  st_transform('ESRI:102241') %>%
  dplyr::select( -NAME, -moe) %>%
  spread(variable, estimate) %>%
  dplyr::select(-geometry) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         Blacks = B02001_003,
         AmInd = B02001_004,
         Asian = B02001_005,
         Hawaiian = B02001_006,
         Hispanic = B03001_003,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         WorkingClassM1517 = B01001_006,
         WorkingClassM1819 = B01001_007,
         WorkingClassM20 = B01001_008,
         WorkingClassM21 = B01001_009,
         WorkingClassM2224 = B01001_010,
         WorkingClassM2529 = B01001_011,
         WorkingClassM3034 = B01001_012,
         WorkingClassM3539 = B01001_013,
         WorkingClassM4044 = B01001_014,
         WorkingClassM4549 = B01001_015,
         WorkingClassM5054 = B01001_016,
         WorkingClassF1517 = B01001_030,
         WorkingClassF1819 = B01001_031,
         WorkingClassF20 = B01001_032,
         WorkingClassF21 = B01001_033,
         WorkingClassF2224 = B01001_034,
         WorkingClassF2529 = B01001_035,
         WorkingClassF3034 = B01001_036,
         WorkingClassF3539 = B01001_037,
         WorkingClassF4044 = B01001_038,
         WorkingClassF4549 = B01001_039,
         WorkingClassF5054 = B01001_040,
         MaleMarried = B12001_004,
         FemaleMarried = B12001_013
  ) %>%
  mutate(pctWorking = ifelse(TotalPop > 0, ((WorkingClassM1517 + WorkingClassM1819 
                                             + WorkingClassM20 + WorkingClassM21 
                                             + WorkingClassM2224 + WorkingClassM2529 
                                             + WorkingClassM3034 + WorkingClassM3539 
                                             + WorkingClassM4044 + WorkingClassM4549 
                                             + WorkingClassM5054 + WorkingClassF1517 
                                             + WorkingClassF1819 + WorkingClassF20 
                                             + WorkingClassF21   + WorkingClassF2224 
                                             + WorkingClassF2529 + WorkingClassF3034 
                                             + WorkingClassF3539 + WorkingClassF4044 
                                             + WorkingClassF4549 + WorkingClassF5054)
                                            /TotalPop * 100), 0),
         pctMarried = ifelse(TotalPop > 0, ((FemaleMarried + MaleMarried) / TotalPop * 100), 0),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop *100, 0),
         pctOtherRace = ifelse(TotalPop > 0, ((Blacks + AmInd +Asian
                                               + Hawaiian + Hispanic) / TotalPop * 100), 0),
         year = "2017") %>%
  dplyr::select(-Whites,-FemaleMarried,-MaleMarried,-Blacks,-AmInd,-Asian,-Hawaiian,
                -Hispanic,-WorkingClassM1517,-WorkingClassM1819,-WorkingClassM20,
                -WorkingClassM21,-WorkingClassM2224,-WorkingClassM2529, 
                -WorkingClassM3034,-WorkingClassM3539,-WorkingClassM4044,
                -WorkingClassM4549,-WorkingClassM5054,-WorkingClassF1517, 
                -WorkingClassF1819,-WorkingClassF20,-WorkingClassF21,-WorkingClassF2224,
                -WorkingClassF2529,-WorkingClassF3034,
                -WorkingClassF3539,-WorkingClassF4044,-WorkingClassF4549,-WorkingClassF5054)

#----- Finding max values of popluation and rent to normalise the data 
# as the tract is not truly representative of the data near the station------

maxtotpop_2017 <- max(tracts217$TotalPop)
maxrent_2017 <- max(tracts217$MedRent , na.rm = T)
tracts2017 <- 
  tracts217 %>%
  mutate( normpop = TotalPop/maxtotpop_2017,
          normrent = MedRent/maxrent_2017)
```

## Combing the year 2009 and 2017 tracts

```{r combine_data for year 2009 and 2017, message = FALSE}
finalTract <- rbind(tracts2009,tracts2017)
```

## Data wrangling with BART data

The BART data we had was in a csv with no coordinate column hence we first separated that string value to get the X and Y for latitude and longitude on which the data could be plotted on the tract data.

```{r Bart data, message = FALSE}
x <- vector(mode='list', length = 105)
y <- vector(mode='list', length = 105)
spli <- strsplit(sfo$Location, ",")
for (val in spli){
  x <- append(x, val[[1]])
  y <- append(y, val[[2]])
  
}
sfo_new <- 
  sfo %>%
  mutate(X = x[106:210],
         Y = y[106:210]) %>%
  dplyr::select(-Abbreviation, -Location, -Description)
crs_tract09 <- st_crs(tracts2009)
sfo_spatial <- st_as_sf(sfo_new, coords = c("X","Y"), crs = 4326, agr = "constant")
sfo_spatial <- sfo_spatial %>% st_transform(st_crs(tracts2009))

# Plotting the lines on the bay area tracts

ggplot() + 
  geom_sf(data=st_union(tracts2009), colour = '#efefef') +
  geom_sf(data=sfo_spatial, 
          aes(colour = Line_colour), 
          show.legend = "point", size= 1.2) +
  labs(title="Bart Stops", 
       subtitle="Bay Area, CA", 
       caption="Figure 1.0") +
  mapTheme()
```

## Creating the buffer zone 

Now that we have our BART and tract data lets get on which creating a buffer from each stop of **0.5 miles**. Note that in the code we use **2640** as the crs units is in feet and we have converted that distance.

```{r Buffer Zone, message = FALSE, warning=FALSE}

# Both a buffer for each stop, and a union of the buffers...
# and bind these objects together

bartBuffers <- 
  rbind(
    st_buffer(sfo_spatial, 2640) %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
    st_union(st_buffer(sfo_spatial, 2640)) %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))

# Let's examine both buffers by making a small multiple
# "facet_wrap" plot showing each

ggplot() +
  geom_sf(data=bartBuffers) +
  geom_sf(data=sfo_spatial, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 1.1") +
  mapTheme()
```

## Finding tracts which lie within the buffer zone

To do this, we clip the union buffer with the tracts, and find the tracts within this intersection. As we do not have the census distribution within the tracts, will we know not which attribute of the census belongs within 0.5 mile radius of each transit station. So to over come this challenge with data, we assume that if the centroid of the census tract lies within the buffer zone, its attributes will be significant.

```{r Buffer Zone_clip, message = FALSE, warning=FALSE}
# Create an sf object with ONLY the unioned buffer
buffer <- filter(bartBuffers, Legend=="Unioned Buffer")

# Clip the 2009 tracts ... by seeing which tracts intersect (st_intersection)
# with the buffer and clipping out only those areas
clip <- 
  st_intersection(buffer, tracts2009) %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Clip")

ggplot() +
  geom_sf(data = clip) +
  labs(title="Clipped Tracts", 
         subtitle="Bay Area, CA", 
         caption="Figure 2.0") +
  theme(plot.title = element_text(size=22))

# Do a spatial selection to see which tracts touch the buffer
selection <- 
  tracts2009[buffer,] %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Selection")

ggplot() +
  geom_sf(data = selection) +
  geom_sf(data = buffer, fill = "transparent", color = "red")+
  labs(title="Selected Tracts", 
         subtitle="Bay Area, CA", 
         caption="Figure 2.1") +
  theme(plot.title = element_text(size=22))

# Do a centroid-in-polygon join to see which tracts have their centroid in the buffer
# Note the st_centroid call creating centroids for each feature
selectCentroids <-
  st_centroid(tracts2009)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(tracts2009, GEOID)) %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")

ggplot() +
  geom_sf(data = selectCentroids) +
  geom_sf(data = buffer, fill = "transparent", color = "red")+
  labs(title="Selected Tracts with centroids in the buffer zone", 
       subtitle="Bay Area, CA", 
       caption="Figure 2.2") +
  theme(plot.title = element_text(size=22))
```

